### 2018-03-12 paper/3-compute.tex

\noindent
\hs{Compute} represents an abstract set of \emph{build tasks}, for example, build
rules in software build systems or formulas in spreadsheets. It is completely
isolated from the world of compilers, file systems, calculation chains, caches,
and all other complexities of real build systems. We explain the compute type
in~\S\ref{sec-compute-applicative}, discussing several noteworthy special cases:
functorial, applicative and monadic types of compute.

@@

\noindent
Since the first argument is polymorphic in \hs{f}, we can choose any \hs{f} that
satisfies the \hs{Applicative} constraint. A good choice in this case is the
applicative functor \hs{Const} defined in the standard module
\cmd{Data.Functor.Const}:

@@

\noindent
A value of type \hs{Const}~\hs{[}\hs{k]}~\hs{v} contains no values \hs{v} but
allows us to record all invocations of the custom \hs{fetch} function that we
feed to the given compute: \hs{fetch k = Const [k]}. The compute returns a
value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)}, which we unwrap
using \hs{maybe [@@] getConst} since inputs have no dependencies.

We can also execute a compute on a given store to obtain the resulting value.
This uses a similar transformation based on the \hs{Identity} monad
defined in \cmd{Data.Functor.Identity}:

@@

...

>>>>

\noindent
\hs{Compute} represents an abstract collection of \emph{build tasks}, for
example, build rules in software build systems or formulas in spreadsheets. It
is completely isolated from the world of compilers, calculation chains, file
systems, caches, and all other complexities of real build systems. We explain
the compute type in~\S\ref{sec-compute-applicative}, discussing several
noteworthy special cases: functorial, applicative and monadic types of compute.

@@

\noindent
Since the first argument is polymorphic in \hs{f}, we can choose any \hs{f} that
satisfies the \hs{Applicative} constraint. A good choice in this case is the
applicative functor \hs{Const} defined in GHC's standard library module
\cmd{Data.Functor.Const}:

@@

\noindent
A value of type \hs{Const}~\hs{[}\hs{k]}~\hs{v} contains no values \hs{v} but
allows us to record all invocations of the custom \hs{fetch} function that we
feed to the given compute: \hs{fetch k = Const [k]}. The compute returns a
value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)}, which we can unwrap
using \hs{maybe [@@] getConst}, i.e. returning the empty list of dependencies
in case of an input.

We can also execute a compute on a given store. This relies on a similar
transformation, now based on the \hs{Identity} monad defined in
\cmd{Data.Functor.Identity}:

@@

As discussed in~\S\ref{sec-background}, some build tasks have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such build tasks correspond to \emph{monadic compute}, i.e.
the type \hs{Compute}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
example:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{bash}
A1 = 10
A2 = 20
B1 = IF(C1=1,A1,A2)
C1 = 1
\end{minted}
\vspace{1mm}

\noindent
The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
our compute abstraction as follows\footnote{The spreadsheet example that uses
the \hs{INDIRECT} function can be expressed as well: simply replace the last
line of the \hs{select} function with \hs{fetch ("A" ++ show c1)}.}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
select :: Compute Monad String Integer
select fetch key | key /= "B1" = Nothing
                 | otherwise   = Just $ do
                     c1 <- fetch "C1"
                     if c1 == 1 then fetch "A1" else fetch "A2"
\end{minted}
\vspace{1mm}

We cannot statically analyse \hs{select} using the function \hs{dependencies}
from the previous section, but we can still \hs{execute} it on a given store:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store key = if key == "A1" then 10 else 20
@\ghci@ execute select store "B1"
Just 20
\end{minted}
\vspace{1mm}

% Note that this presentation of build tasks makes dynamic dependencies
% immediately recognisable: we need to examine the actual value \hs{c1} by
% extracting it from the computational context using \hs{Monad}'s bind statement
% \hs{c1 <- fetch}
